Unified Convergence Analysis of FedAve on Non-IID Data

### 2020-04-19 source/sec_sgd_scvx_smooth.tex

%
\begin{assumption}
The stochastic gradients have bounded variance 
\begin{align*}
\mathbb{E}\|\nabla F_{k}(w_{t}^{k},\xi_{t}^{k})-\nabla F_{k}(w_{t}^{k})\|^{2} & \leq\sigma_{k}^{2}
\end{align*}
for all $k$. Moreover, the expected squared norm of stochastic gradients
is uniformly bounded, 
\begin{align*}
\mathbb{E}\|\nabla F_{k}(w_{t}^{k},\xi_{t}^{k})\|^{2} & \leq G^{2}
\end{align*}
\end{assumption}
\begin{theorem}
(Full device participation) Suppose $F_{k}$ is $L$-smooth and $\mu$-strongly
convex for all $k$. Let $\kappa=\frac{L}{\mu}$, $\gamma=\max\{8\kappa-1,E\}$
where $E$ is the communication interval, and learning rates 
\begin{align*}
\alpha_{t} & =\frac{c}{\mu(\gamma+t)}
\end{align*}
so that $\alpha_{t}\leq2\alpha_{t+E}$, and where $0\leq c\leq1$
is small enough such that the following hold: 
\begin{align*}
\alpha_{t} & \leq\frac{1}{4L}\\
\alpha_{t} & \leq\frac{1}{N}
\end{align*}
 for all $t\geq0$.

Then with full device participation,
\begin{align*}
\mathbb{E}F(w_{T})-F^{\ast} & \leq C\frac{\kappa}{N(\gamma+t)}\\
C & =4E^{2}LG^{2}+\nu_{max}^{2}\sigma^{2}+LG^{2}
\end{align*}

\end{theorem}
\begin{proof}
The idea of the proof is to observe that the $L$-smoothness of $F$
provides the upper bound
\begin{align*}
\mathbb{E}(F(\overline{w}_{t}))-F^{\ast} & =\mathbb{E}(F(\overline{w}_{t})-F(w^{\ast}))\\
 & \leq\frac{L}{2}\mathbb{E}\|\overline{w}_{t}-w^{\ast}\|^{2}
\end{align*}
 and to show that $\mathbb{E}\|\overline{w}_{T}-w^{\ast}\|^{2}=O(\frac{1}{NT})$.

@@

\begin{align*}
\mathbb{E}\|\overline{w}_{t+1}-w^{\ast}\|^{2} & \leq\mathbb{E}(1-\mu\alpha_{t})\|\overline{w}_{t}-w^{\ast}\|^{2}+4E^{2}L\alpha_{t}^{3}G^{2}+4E^{2}L^{2}\alpha_{t}^{4}G^{2}+\alpha_{t}^{2}\sum_{k=1}^{N}p_{k}^{2}\sigma_{k}^{2}+\alpha_{t}^{3}LG^{2}\\
 & =\mathbb{E}(1-\mu\alpha_{t})\|\overline{w}_{t}-w^{\ast}\|^{2}+4E^{2}L\alpha_{t}^{3}G^{2}+4E^{2}L^{2}\alpha_{t}^{4}G^{2}+\alpha_{t}^{2}\frac{1}{N^{2}}\sum_{k=1}^{N}(p_{k}N)^{2}\sigma_{k}^{2}+\alpha_{t}^{3}LG^{2}\\
 & \leq\mathbb{E}(1-\mu\alpha_{t})\|\overline{w}_{t}-w^{\ast}\|^{2}+4E^{2}L\alpha_{t}^{3}G^{2}+4E^{2}L^{2}\alpha_{t}^{4}G^{2}+\alpha_{t}^{2}\frac{1}{N^{2}}\nu_{max}^{2}\sum_{k=1}^{N}\sigma_{k}^{2}+\alpha_{t}^{3}LG^{2}\\
 & \leq\mathbb{E}(1-\mu\alpha_{t})\|\overline{w}_{t}-w^{\ast}\|^{2}+4E^{2}L\alpha_{t}^{3}G^{2}+4E^{2}L^{2}\alpha_{t}^{4}G^{2}+\alpha_{t}^{2}\frac{1}{N}\nu_{max}^{2}\sigma^{2}+\alpha_{t}^{3}LG^{2}
\end{align*}
 where $\sigma^{2}=\max_{k}\sigma_{k}^{2}$. The leading term in the
bound is $\alpha_{t}^{2}\frac{1}{N}\nu_{max}^{2}\sigma^{2}$, which
one summed using the recursive relations gives $O(\alpha_{t})$, but
the constant now is scaled down by $1/N$. To achieve this linear
speedup, we need $\sum_{t=0}^{T}(1-\mu\alpha_{t})\|w_{0}-w^{\ast}\|^{2}$
to be small enough.

@@

Finally, the $L$-smoothness of $F$ implies 
\begin{align*}
\mathbb{E}(F(\overline{w}_{t}))-F^{\ast} & =\mathbb{E}(F(\overline{w}_{t})-F(w^{\ast}))\\
 & \leq\frac{L}{2}\mathbb{E}\|\overline{w}_{t}-w^{\ast}\|^{2}\leq\frac{L}{2}\frac{v}{\gamma+t}
\end{align*}
 Now if we choose $\delta=\frac{c}{\mu}$ where $c\leq1$ is small
enough such that the required conditions 
\begin{align*}
\alpha_{t}^{2}+\beta_{t-1}^{2} & \leq\frac{1}{2}\\
\alpha_{t} & \leq\frac{1}{4L}\\
4\alpha_{t-1}^{2} & \leq\alpha_{t}
\end{align*}
hold for all $t\geq0$, $\gamma=\max\{8\kappa-1,E\}$, then $\alpha_{t}=\frac{c}{\mu}\frac{1}{\gamma+t}$
and recalling $v=\max\{\frac{\delta^{2}B'}{\delta\mu-c+1},(\gamma+1)\|w_{0}-w^{\ast}\|^{2}\}$,
we finally have 
\begin{align*}
\mathbb{E}(F(\overline{w}_{t}))-F^{\ast} & \leq\frac{L}{2}\frac{v}{\gamma+t}\leq\frac{L/2}{\gamma+t}(\frac{\delta^{2}B'}{\delta\mu-c+1}+(\gamma+1)\|w_{0}-w^{\ast}\|^{2})\\
 & \leq\frac{\kappa}{\gamma+t}(\frac{B'}{\mu}+4L\|w_{0}-w^{\ast}\|^{2})
\end{align*}
\end{proof}

>>>>

We first summarize some properties of $L$-smooth and $\mu$-strongly
convex functions.

\begin{assumption}
	The stochastic gradients have bounded variance 
	\begin{align*}
	\mathbb{E}\|\nabla F_{k}(w_{t}^{k},\xi_{t}^{k})-\nabla F_{k}(w_{t}^{k})\|^{2} & \leq\sigma_{k}^{2}
	\end{align*}
	for all $k$. Moreover, the expected squared norm of stochastic gradients
	is uniformly bounded, 
	\begin{align*}
	\mathbb{E}\|\nabla F_{k}(w_{t}^{k},\xi_{t}^{k})\|^{2} & \leq G^{2}
	\end{align*}
\end{assumption}
\begin{theorem}
	Suppose $F_{k}$ is $L$-smooth and $\mu$-strongly convex for all
	$k$. Let $\kappa=\frac{L}{\mu}$, $\gamma=\max\{8\kappa-1,E\}$ where
	$E$ is the communication interval, and learning rates 
	\begin{align*}
	\alpha_{t} & =\frac{c}{\mu(\gamma+t)}
	\end{align*}
	so that $\alpha_{t}\leq2\alpha_{t+E}$, and where $0\leq c\leq1$
	is small enough such that the following hold: 
	\begin{align*}
	\alpha_{t} & \leq\frac{1}{4L}\\
	\alpha_{t} & \leq\frac{1}{N}
	\end{align*}
	for all $t\geq0$.

	Then with full device participation, as long as $E\leq O(\sqrt{\frac{N}{T}})$,
	\begin{align*}
	\mathbb{E}F(w_{T})-F^{\ast} & \leq C\frac{\kappa}{N(\gamma+t)}\\
	C & =5LG^{2}+\nu_{max}^{2}\sigma^{2}
	\end{align*}
	and with partial participation, with $E=O(1)$ and $K$ sampled devices
	at each communication round, 
	\begin{align*}
	\mathbb{E}F(w_{T})-F^{\ast} & \leq(4E^{2}G^{2}+\nu_{max}^{2}\sigma^{2})\frac{\kappa}{K(\gamma+t)}+\frac{4E^{2}LG^{2}}{t^{2}}
	\end{align*}
\end{theorem}
\begin{proof}
	The idea of the proof is to observe that the $L$-smoothness of $F$
	provides the upper bound
	\begin{align*}
	\mathbb{E}(F(\overline{w}_{t}))-F^{\ast} & =\mathbb{E}(F(\overline{w}_{t})-F(w^{\ast}))\\
	& \leq\frac{L}{2}\mathbb{E}\|\overline{w}_{t}-w^{\ast}\|^{2}
	\end{align*}
	and to show that $\mathbb{E}\|\overline{w}_{T}-w^{\ast}\|^{2}=O(\frac{1}{NT})$.

@@

	\begin{align*}
	\mathbb{E}\|\overline{w}_{t+1}-w^{\ast}\|^{2} & \leq\mathbb{E}(1-\mu\alpha_{t})\|\overline{w}_{t}-w^{\ast}\|^{2}+4E^{2}L\alpha_{t}^{3}G^{2}+4E^{2}L^{2}\alpha_{t}^{4}G^{2}+\alpha_{t}^{2}\sum_{k=1}^{N}p_{k}^{2}\sigma_{k}^{2}+\alpha_{t}^{3}LG^{2}\\
	& =\mathbb{E}(1-\mu\alpha_{t})\|\overline{w}_{t}-w^{\ast}\|^{2}+4E^{2}L\alpha_{t}^{3}G^{2}+4E^{2}L^{2}\alpha_{t}^{4}G^{2}+\alpha_{t}^{2}\frac{1}{N^{2}}\sum_{k=1}^{N}(p_{k}N)^{2}\sigma_{k}^{2}+\alpha_{t}^{3}LG^{2}\\
	& \leq\mathbb{E}(1-\mu\alpha_{t})\|\overline{w}_{t}-w^{\ast}\|^{2}+4E^{2}L\alpha_{t}^{3}G^{2}+4E^{2}L^{2}\alpha_{t}^{4}G^{2}+\alpha_{t}^{2}\frac{1}{N^{2}}\nu_{max}^{2}\sum_{k=1}^{N}\sigma_{k}^{2}+\alpha_{t}^{3}LG^{2}\\
	& \leq\mathbb{E}(1-\mu\alpha_{t})\|\overline{w}_{t}-w^{\ast}\|^{2}+4E^{2}L\alpha_{t}^{3}G^{2}+4E^{2}L^{2}\alpha_{t}^{4}G^{2}+\alpha_{t}^{2}\frac{1}{N}\nu_{max}^{2}\sigma^{2}+\alpha_{t}^{3}LG^{2}
	\end{align*}
	where $\sigma^{2}=\max_{k}\sigma_{k}^{2}$. The leading term in the
	bound is $\alpha_{t}^{2}\frac{1}{N}\nu_{max}^{2}\sigma^{2}$, which
	one summed using the recursive relations gives $O(\alpha_{t})$, but
	the constant now is scaled down by $1/N$. To achieve this linear
	speedup, we need $\sum_{t=0}^{T}(1-\mu\alpha_{t})\|w_{0}-w^{\ast}\|^{2}$
	to be small enough. \textbf{Note also that the above bound implies
		that $E$ can be chosen to be as large as $O(\sqrt{\frac{T}{N}})$
		without degrading the performance, since $E^{2}\alpha_{t}^{3}=O(\alpha_{t}^{2}/N)$
		which is the order of the leading term. As we will see, in the partial
		participation case, this is no longer true, and $E$ cannot be chosen
		$O(T^{\beta})$ for any $\beta>0$. }

@@

	Finally, the $L$-smoothness of $F$ implies 
	\begin{align*}
	\mathbb{E}(F(\overline{w}_{t}))-F^{\ast} & =\mathbb{E}(F(\overline{w}_{t})-F(w^{\ast}))\\
	& \leq\frac{L}{2}\mathbb{E}\|\overline{w}_{t}-w^{\ast}\|^{2}\leq\frac{L}{2}\frac{v}{\gamma+t}
	\end{align*}
	Now if we choose $\delta=\frac{c}{\mu}$ where $c\leq1$ is small
	enough such that the required conditions 
	\begin{align*}
	\alpha_{t}^{2}+\beta_{t-1}^{2} & \leq\frac{1}{2}\\
	\alpha_{t} & \leq\frac{1}{4L}\\
	4\alpha_{t-1}^{2} & \leq\alpha_{t}
	\end{align*}
	hold for all $t\geq0$, $\gamma=\max\{8\kappa-1,E\}$, then $\alpha_{t}=\frac{c}{\mu}\frac{1}{\gamma+t}$
	and recalling $v=\max\{\frac{\delta^{2}B'}{\delta\mu-c+1},(\gamma+1)\|w_{0}-w^{\ast}\|^{2}\}$,
	we finally have 
	\begin{align*}
	\mathbb{E}(F(\overline{w}_{t}))-F^{\ast} & \leq\frac{L}{2}\frac{v}{\gamma+t}\leq\frac{L/2}{\gamma+t}(\frac{\delta^{2}B'}{\delta\mu-c+1}+(\gamma+1)\|w_{0}-w^{\ast}\|^{2})\\
	& \leq\frac{\kappa}{\gamma+t}(\frac{B'}{\mu}+4L\|w_{0}-w^{\ast}\|^{2})
	\end{align*}

	With partial participation, the update at each communication round
	is now given by averages over a subset of sampled devices. The key
	is to bound 
	\begin{align*}
	\mathbb{E}\|\overline{w}_{t}-\overline{y}_{t}\|^{2} & =\frac{1}{K}\sum_{k}p^{k}\|w_{t}^{k}-\sum_{k}p_{k}w_{t}^{k}\|^{2}
	\end{align*}
	This can be bounded by the same argument as before and yield 
	\begin{align*}
	\mathbb{E}\|\overline{w}_{t}-\overline{y}_{t}\|^{2} & \leq\frac{4}{K}\alpha_{t}^{2}E^{2}G^{2}
	\end{align*}
	and so this is also a leading term, and together with the term $\alpha_{t}^{2}\frac{1}{N}\nu_{max}^{2}\sigma^{2}$
	determines the convergence rate, which is now $O(\frac{(\nu_{max}^{2}\sigma^{2}+4E^{2}G^{2})}{KT})$.
	This implies that $E$ canot be chosen $O(T^{\beta})$ for any $\beta>0$
	without degrading the performance. This should be checked in experiments,
	whether with partial participation if the communication round is set
	to scale with $T$, the convergence deteriorates. This is in constrast
	with the full participation case, where $E=O(\sqrt{\frac{T}{N}})$
	is allowed. \textbf{If we can confirm that full participation allows
		linear speedup with $\nu=N\cdot\max_{k}p_{k}\approx1$ and $E=O(\sqrt{\frac{T}{N}})$,
		whereas partial participation only allows $E=O(1)$, then this would
		an interesting phenomenon that is not reported by previous studies!}
\end{proof}
